Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study

### 2015-01-02 chemotaxis/paper/draft1.tex
edited paper

\section{Issues in Manifold Learning/Diffusion maps}

\subsection{Diffusion maps}

@@

\subsection{Observers}

@@

\subsection{Metric}

@@

\section{Identifying the Principal Components}

@@

Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^n$, we would like to automatically deduce which ones are harmonics/functions of another. 
%
To accomplish this, we will fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
%
If the resulting fit is good/accurate, then $\phi_{k}$ is most likely a harmonic of $\phi_1, \dots, \phi_{k-1}$.

We will use a local linear function 
\begin{equation}
\phi_k \approx \alpha( \Phi_{k-1}) + \beta(\Phi_{k-1})^T \Phi_{k-1}
\end{equation}
where 
%
$\Phi_{k-1} = \begin{bmatrix} \phi_1 \\ \vdots \\ \phi_{k-1} \end{bmatrix}$
and $\beta \in \mathbb{R}^{k-1}$. 
%
The coefficients $\alpha$ and $\beta$ are {\em functions} of $\phi_{k-1}$ because we use a {\em local} linear fit; therefore, the coefficients change as a function of the domain.

At each point $\Phi_{k-1}(i)$, we fit a local linear function using the remaining $m-1$ data points. 
%
We solve the following optimization problem 
\begin{equation} \label{eq:opt_problem}
\hat{\alpha} (i) , \hat{\beta}(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} w_i(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
\end{equation}
%
where $w$ is a weighting function.
%
We will use Gaussian weights, 
%
\begin{equation}
w_i(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon^2} \right).
\end{equation}
%
We take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$.

@@

Our first illustrative example will consist of two different swiss roll data sets.
%
Each data set consists of $m=1500$ points uniformly distributed on the swiss roll. 
%
Both swiss rolls are the same in the angular/spiral direction.
%
However, height of the first swiss roll is $h = 15 \pi$, while the heigh of the second is $h = 6 \pi$. 
%
These two data sets are shown in Figure~\ref{fig:swiss_rolls}(a)~and~(b), respectively.
%
We expect to recover the eigenvector which parameterizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.

@@

@@

\begin{figure}[t]
%
\centering
\includegraphics[width=\textwidth]{chemotaxis_compare_timescales_evals}
%
\caption{}
%
\label{fig:chemotaxis_compare_timescales_evals}
%
\end{figure}

>>>>

\section{Background on Manifold Learning and Some Relevant Issues}

\subsection{Manifold learning algorithm: diffusion maps}

@@

\subsection{Constructing appropriate observers}

@@

\subsection{Defining a meaningful metric}

@@

\section{Identifying the Informative Eigenvectors}

@@

Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^n$, we would like to automatically deduce which ones are harmonics/functions of another. 
%
To accomplish this, we will fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
%
If the resulting fit is good/accurate, then $\phi_{k}$ is most likely a harmonic of $\phi_1, \dots, \phi_{k-1}$. 
%
We will use a local linear function 
\begin{equation}
\phi_k \approx \alpha( \Phi_{k-1}) + \beta(\Phi_{k-1})^T \Phi_{k-1}
\end{equation}
where 
%
$\Phi_{k-1} = \begin{bmatrix} \phi_1 \\ \vdots \\ \phi_{k-1} \end{bmatrix}$
and $\beta \in \mathbb{R}^{k-1}$. 
%
The coefficients $\alpha$ and $\beta$ are functions of $\phi_{k-1}$ because we use a {\em local} linear fit; therefore, the coefficients change as a function of the domain.

At each point $\Phi_{k-1}(i)$, we approximate $\phi_k(i)$ by fitting a local linear function using the remaining $m-1$ data points. 
%
We solve the following optimization problem 
\begin{equation} \label{eq:opt_problem}
\hat{\alpha} (i) , \hat{\beta}(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} w_i(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
\end{equation}
%
where $w$ is a weighting function.
%
We will use Gaussian weights, 
%
\begin{equation}
w_i(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon^2} \right).
\end{equation}
%
We take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$.
%
The solution of \eqref{eq:opt_problem} is given by the solution of an eigenvector problem \cite{...}.

@@

We will demonstrate the local linear regression algorithm for identifying informative eigenvectors on two synthetic data sets. 
%
Our first illustrative example will consist of two different swiss roll data sets.
%
Each data set consists of $m=1500$ points uniformly distributed on the swiss roll. 
%
Both swiss rolls are the same in the angular/spiral direction.
%
However, height of the first swiss roll is $h = 15 \pi$, while the heigh of the second is $h = 6 \pi$. 
%
These two data sets are shown in Figure~\ref{fig:swiss_rolls}(a)~and~(b), respectively.
%
We expect to recover the eigenvector which parameterizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.

@@

The issues discussed above must all be considered and addressed in order to successfully analyze data using manifold learning algorithms. 
%
To illustrate this point, we will use a model problem arising in the study of cellular dynamics. 
%
In this example, constructing appropriate observers with an informative metric will allow us to perform informative analysis. 
%
Furthermore, automatically detecting the ``important'' components will allow us to deduce when the dimension of the underlying dynamical system changes. 
%
In this way, we can deduce how many variables are required to write an accurate reduced model.

@@

\begin{figure}[t]
%
\centering
\includegraphics[width=\textwidth]{chemotaxis_compare_timescales_evals}
%
\caption{Detecting the change in dimensionality. The ratio of the log of the two ``meaningful'' diffusion maps eigenvalues (those which correspond to modes which are not harmonics) is shown in blue, and the ratio of the diffusive and drift time scales are shown in red. When the diffusive time scale is significantly larger than the drift time scale, the data is expected to be effectively one-dimensional. This transition is consistent with the diffusion maps eigenvalues: the data is effectively one-dimensional if the first eigenvalue is significantly larger than the second. }
%
\label{fig:chemotaxis_compare_timescales_evals}
%
\end{figure}
